---
title: "Assignment 2"
author: "Your name"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    theme: united
    df_print: paged
    highlight: textmate
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: inline
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Objectives

- Manipulating Data
- Categorizing Data
- Sorting and Grouping Data
- Summarizing Data
- Combining Data
- NestedQueries
- Views and Indexes
- Transforming Data

Note: Bonus points given for answering optional questions, proper formatting, and comments.

__SIGNIFICANT BONUS__: If you are feeling ambitious, you have the option of answering the data manipulation language (DML) questions in dplyr/dbplyr (in addition to already answering the question in SQL). Data Definition Language (DDL) questions are not a part of this bonus opportunity (`CREATE`, `ALTER`, `DROP`, `TRUNCATE`, `COMMENT`, `RENAME`).

```{r message=FALSE, warning=FALSE}
# If you need these packages installed, uncomment the line of code and install them
# install.packages("RMySQL")
# install.packages("keyring")
# install.packages("odbc")
# install.packages("tidyverse")
#library(keyring)
library(RMySQL)
library(odbc)
library(tidyverse)
```

```{r}
# Give your MySQL a service name or specify a service name for a key already created
# I chose r-mysql as the service name but you can change it
service_mysql <- "r-mysql"
# Your username should be root unless you changed it during MySQL installation
username_mysql <- "root"
```

```{r eval=FALSE}
# THIS CODE CHUNK IS ONLY RAN ONCE INTERACTIVE; see week-3-introducing-r-with-sql file for more information
# We set this chunk to eval=FALSE so when we knit
# So R ignores this chunk (we only have to make a key once)
# Create a keyring to access MySQL if you have not yet
#keyring::key_set(service = service_mysql, username = username_mysql)
```

```{r}
# Change this if you used a different name for service
# Connect to database
con_sakila <- dbConnect(
  MySQL(),
  user = username_mysql,
  password = rstudioapi::askForPassword("Database password"),
  dbname = "sakila",
  host = "localhost"
)
```

Using the sakila database, provide the SQL queries you used. If there is more than one solution, provide only one solution. 

# QUESTION 1

### a) Show the list of databases.

```{sql connection=con_sakila}
SHOW DATABASES;
```

### b) Select sakila database.

```{sql connection=con_sakila}
USE sakila
```

### c) Show all tables in the sakila database.

```{sql connection=con_sakila}
SHOW TABLES
```

### d) Show each of the columns along with their data types for the actor table.

```{sql connection=con_sakila}
DESCRIBE actor
```

### e) Show the total number of records in the actor table. (dplyr bonus)

```{sql connection=con_sakila}
SELECT COUNT(*)
FROM actor
```

### e.2) dplyr bonus

```{r}
actor_table <- DBI::dbReadTable(con_sakila, "actor")
count(actor_table)
```

### f) What is the first name and last name of all the actors in the actor table? (dplyr bonus)

```{sql connection=con_sakila}
SELECT actor_id, first_name, last_name FROM ACTOR
```
### f.2) dplyr bonus

```{r}
actor_table %>% select(actor_id, first_name, last_name)
```

### g) Insert your first name and middle initial (in the last name column) into the actors table.

```{sql connection=con_sakila}
INSERT INTO actor (first_name, last_name) VALUES ('Stathy', 'G')
```

```{sql connection=con_sakila}
SELECT first_name, last_name FROM actor ORDER BY last_update DESC LIMIT 1
```

### h) Update your middle initial with your last name in the actors table.

```{sql connection=con_sakila}
UPDATE actor SET last_name = 'Touloumis' WHERE first_name = 'Stathy'
```

```{sql connection=con_sakila}
SELECT COUNT(actor_id) FROM actor WHERE first_name = 'Stathy' AND last_name = 'Touloumis'
```

### i) Delete the record from the actor table where the first namematches your first name.

```{sql connection=con_sakila}
DELETE FROM actor WHERE first_name = 'Stathy'
```

```{sql connection=con_sakila}
SELECT COUNT(actor_id) FROM actor WHERE first_name = 'Stathy'
```

### j) Create a table payment_type with the following specifications and appropriate data types

Table Name : 'payment_type'
Primary Key: 'payment_type_id'
Column: 'Type'
Insert following rows in to the table: 1, 'Credit Card'; 2, 'Cash'; 3, 'Paypal'; 4 , 'Cheque'

```{sql connection=con_sakila}
CREATE TABLE payment_type (
  payment_type_id INT NOT NULL AUTO_INCREMENT,
  type VARCHAR(12) NOT NULL,
  PRIMARY KEY(payment_type_id)
)
```

```{sql connection=con_sakila}
INSERT INTO payment_type (type) VALUES ('Credit Card')
```

```{sql connection=con_sakila}
INSERT INTO payment_type (type) VALUES ('Cash')
```

```{sql connection=con_sakila}
INSERT INTO payment_type (type) VALUES ('Paypal')
```

```{sql connection=con_sakila}
INSERT INTO payment_type (type) VALUES ('Cheque')
```

### k) Rename table payment_type to payment_types.

```{sql connection=con_sakila}
RENAME TABLE payment_type TO payment_types
```

### l) Drop the table payment_types.

```{sql connection=con_sakila}
DROP TABLE payment_types
```

# QUESTION 2

# a) List all the movies (title & description) that are rated PG-13. (dplyr bonus)

```{sql connection=con_sakila}
SELECT title, description FROM film WHERE rating = 'PG-13'
```
### a.2) dplyr bonus

```{r}
film_tbl <- DBI::dbReadTable(con_sakila, "film")
film_tbl %>%
  filter(rating == 'PG-13') %>%
  select(title, description)
```

# b) List all movies that are either PG OR PG-13 using `IN` operator. (dplyr bonus)

```{sql connection=con_sakila}
SELECT title, description FROM film WHERE rating IN ( 'PG-13', 'PG' )
```

# c) Report all payments greater than and equal to $2 and Less than equal to $7. (dplyr bonus)

```{sql connection=con_sakila}
SELECT payment_id, payment_date, amount FROM payment WHERE amount >= 2.00 AND amount <= 7.00
```

Note : write 2 separate queries conditional operator and `BETWEEN` keyword

### d) List all addresses that have phone number that contain digits 589, start with 140 or end with 589. (dplyr bonus)

```{sql connection=con_sakila}
SELECT * FROM address
WHERE phone LIKE '%589%'
  OR phone LIKE '140%'
  OR phone LIKE '%589'
```

Note : write 3 different queries

### e) List all staff members (first name, last name, email) who have no password. (dplyr bonus)

```{sql connection=con_sakila}
SELECT first_name, last_name, email
FROM staff
WHERE password IS NULL
```

```{sql connection=con_sakila}
SELECT first_name, last_name, email
FROM staff
WHERE password = ''
  OR password IS NULL
```

### f) Select all films that have title names like ZOO and rental duration greater than or equal to 4. (dplyr bonus)

```{sql connection=con_sakila}
SELECT *
FROM film
WHERE title LIKE '%ZOO%'
  AND rental_duration >= 4
```

### g) What is the cost of renting the movie ACADEMY DINOSAUR for 2 weeks?

```{sql connection=con_sakila}
SELECT (rental_rate * (14 / rental_duration)) AS 'cost for 2 weeks'
FROM film
WHERE title LIKE '%ACADEMY DINOSAUR%'
```

Note : use of column alias

### h) List all unique districts where the customers, staff, and stores are located. (dplyr bonus)

```{sql connection=con_sakila}
SELECT DISTINCT district 'Unique Districts'
FROM address
```

Note : check for `NOT NULL` values

### i) List the top 10 newest customers across all stores. (dplyr bonus)

```{sql connection=con_sakila}
SELECT CONCAT(first_name, ' ', last_name) AS 'CUSTOMER'
FROM customer
ORDER BY create_date DESC
  LIMIT 10
```

# QUESTION 3

### a) Show total number of movies. (dplyr bonus)

```{sql connection=con_sakila}
SELECT COUNT(film_id)
FROM film
```

### b) What is the minimum payment received and max payment received across all transactions? (dplyr bonus)

```{sql connection=con_sakila}
SELECT MIN(amount) AS 'Min Payment', MAX(amount) AS 'Max Payment'
FROM payment
```

### c) Number of customers that rented movies between Feb-2005 and May-2005 (based on payment date). (dplyr bonus)

```{sql connection=con_sakila}
SELECT COUNT(customer_id)
FROM payment
WHERE payment_date BETWEEN '2005-02-01' and '2005-05-31'
```

### d) List all movies where replacement_cost is greater than $15 or rental_duration is between 6 and 10 days. (dplyr bonus)

```{sql connection=con_sakila}
SELECT *
FROM film
WHERE replacement_cost > 15.0
  AND rental_duration BETWEEN 6 AND 10
```

### e) What is the total amount spent by customers for movies in the year 2005? (dplyr bonus)

```{sql connection=con_sakila}
SELECT SUM(amount) 'Total Movie Spend In 2005'
FROM payment
WHERE payment_date BETWEEN '2005-01-01' AND '2005-12-31'
```

### f) What is the average replacement cost across all movies? (dplyr bonus)

```{sql connection=con_sakila}
SELECT AVG(replacement_cost) AS 'Avg Cost To Replace'
FROM film
```

### g) What is the standard deviation of rental rate across all movies? (dplyr bonus)

```{sql connection=con_sakila}
SELECT STD(rental_rate) AS 'Std Deviation Rate'
FROM film
```

### h) What is the midrange of the rental duration for all movies. (dplyr bonus)

```{sql connection=con_sakila}
SELECT (SUM(rental_duration) / COUNT(rental_duration)) AS 'Duration Midrange'
FROM film
```

# QUESTION 4

### a) List all customers that live in the Nepal. (dplyr bonus)

```{sql connection=con_sakila}
SELECT c.*
FROM customer c
INNER JOIN address a
  ON c.address_id = a.address_id
INNER JOIN city ci
  ON a.city_id = ci.city_id
INNER JOIN country co
  ON ci.country_id = co.country_id
WHERE co.country LIKE '%nepal%'
  AND c.active IS TRUE
```

### b) List all actors that appear in the movie titled Academy Dinosaur. (dplyr bonus)

```{sql connection=con_sakila}
SELECT CONCAT(a.first_name, ' ', a.last_name) AS 'Actor Name'
FROM actor a
INNER JOIN film_actor fa
  ON a.actor_id = fa.actor_id
INNER JOIN film f 
  ON f.film_id = fa.film_id
WHERE f.title LIKE '%academy dinosaur%'
```

### c) What is the revenue generated by each customer? (dplyr bonus)

```{sql connection=con_sakila}
SELECT c.customer_id, CONCAT('$', SUM(p.amount)) AS 'Total $'
FROM payment p
INNER JOIN customer c
  ON p.customer_id = c.customer_id
GROUP BY c.customer_id
```

### d) List top 10 customers that rented the most movies.

```{sql connection=con_sakila}
SELECT CONCAT(c.first_name, ' ', c.last_name) 'Top Customer', COUNT(rental_id) AS rental_count
FROM rental r
INNER JOIN customer c
  ON r.customer_id = c.customer_id
GROUP BY c.customer_id
ORDER BY rental_count DESC
  LIMIT 10
```

### e) List the inventory available in store to rent for each of the movies. (dplyr bonus)

```{sql connection=con_sakila}
SELECT f.*
FROM film f
INNER JOIN inventory i
  ON f.film_id = i.film_id
INNER JOIN store s
  ON i.store_id = s.store_id
INNER JOIN rental r
  ON r.inventory_id = i.inventory_id
GROUP BY f.film_id
```

### f) List the top zipcodes that have the highest rental activity. (dplyr bonus)

```{sql connection=con_sakila}
SELECT a.postal_code, COUNT(r.rental_id) AS rental_activity
FROM address a
INNER JOIN store s
  ON s.address_id = a.address_id
INNER JOIN staff st
  ON st.staff_id = a.address_id
INNER JOIN rental r
  ON r.staff_id = st.staff_id
GROUP BY a.postal_code
ORDER BY rental_activity DESC
```
# QUESTION 5

Note: For questions a, b, c below use a single query with a sub query

### a) List actors and customers whose first name is the same as the first name of the actor with ID 8. (dplyr bonus)

```{sql connection=con_sakila}
SELECT CONCAT(c.first_name, ' ', c.last_name) 'Customer Name', CONCAT(a.first_name, ' ',  a.last_name) 'Actor Name'
FROM customer c
INNER JOIN actor a
  ON c.first_name = a.first_name
WHERE a.first_name = (SELECT first_name FROM actor WHERE actor_id = 8)
```
### b) List customers and payment amounts, with payments greater than average payment amount. (dplyr bonus)

```{sql connection=con_sakila}
SELECT c.*, p.amount
FROM customer c
INNER JOIN payment p
  ON p.customer_id = c.customer_id
WHERE p.amount > (SELECT AVG(amount) FROM payment)
```
### c) List customers who have rented movies at least once. (dplyr bonus)

```{sql connection=con_sakila}
SELECT CONCAT(c.first_name, ' ', c.last_name) AS customer, COUNT(r.rental_id) AS rental_cnt
FROM customer c
INNER JOIN rental r
  On r.customer_id = c.customer_id
GROUP BY customer
ORDER BY rental_cnt DESC
```

Note: use `IN` clause with the sub query

### d) Find the floor of the maximum, minimum and average payment amount

```{sql connection=con_sakila}
SELECT FLOOR(MAX(p.amount))
FROM (
SELECT MAX(amount) amount FROM payment
UNION
SELECT MIN(amount) amount FROM payment
UNION
SELECT AVG(amount) amount FROM payment ) AS p
```

# QUESTION 6

### a) Create a view called actors_portfolio which contains information about actors and films ( including titles and category).

```{sql connection=con_sakila}
CREATE VIEW sakila.actors_portfolio AS
  SELECT f.title "film_title", c.name "category"
  FROM actor a
  INNER JOIN film_actor fa
    ON fa.actor_id = a.actor_id
  INNER JOIN film f
    ON f.film_id = fa.film_id
  INNER JOIN film_category fc
    ON fc.film_id = f.film_id
  INNER JOIN category c
    ON c.category_id = fc.category_id 
```

### b) Describe the structure of the view and query the view to get information on the actor ADAM GRANT

```{sql connection=con_sakila}
DESCRIBE actors_portfolio ;
```

```{sql connection=con_sakila}
SELECT DISTINCT * FROM actors_portfolio
```
```{sql connection=con_sakila}
DROP VIEW actors_portfolio
```

### c) Insert a new movie titled Data Hero in Sci-Fi Category starring ADAM GRANT

_You'll be graded for the sql code used to "insert" data instead of the error message._
```{sql connection=con_sakila}
INSERT INTO film (title, language_id)
VALUES(
  'Data Hero',
  (SELECT language_id FROM language WHERE name = 'English') )
```
```{sql connection=con_sakila}
INSERT INTO film_category (film_id, category_id)
VALUES (
  (SELECT film_id FROM film WHERE title = 'Data Hero'),
  (SELECT category_id FROM category WHERE name = 'Sci-Fi')
)
```
```{sql connection=con_sakila}
INSERT INTO film_actor (film_id, actor_id)
VALUES (
  (SELECT film_id FROM film WHERE title = 'Data Hero'),
  (SELECT actor_id FROM actor WHERE first_name = 'ADAM' AND last_name =  'GRANT')
)
```



```{sql connection=con_sakila}
SELECT f.film_id, title, c.name, CONCAT(a.first_name, ' ', a.last_name) "Actor"
FROM film f
INNER JOIN film_category fc
  ON fc.film_id = f.film_id
INNER JOIN category c
  ON c.category_id = fc.category_id
INNER JOIN film_actor fa
  ON fa.film_id = f.film_id
INNER JOIN actor a
  ON a.actor_id = fa.actor_id
WHERE f.title = 'Data Hero'
```

Note: If you see an error, explain why this is not permitted

```{sql connection=con_sakila}
DELETE FROM film_category WHERE film_id IN (SELECT film_id FROM film WHERE title = 'Data Hero')
```
```{sql connection=con_sakila}
DELETE FROM film_actor WHERE film_id IN (SELECT film_id FROM film WHERE title = 'Data Hero')
```
```{sql connection=con_sakila}
DELETE FROM film WHERE title = 'Data Hero'
```

# QUESTION 7 (Optional Practice Questions)

### a) Customers sorted by first name and last name in ascending order. (dplyr bonus)

```{sql connection=con_sakila}
SELECT *
FROM customer
ORDER BY first_name ASC, last_name ASC
```

### b) Group distinct addresses by district. (dplyr bonus)

```{sql connection=con_sakila}
SELECT 1 FROM dual
```

### c) Count of movies that are either G/NC-17/PG-13/PG/R grouped by rating. (dplyr bonus)

```{sql connection=con_sakila}
SELECT COUNT(f.film_id)
FROM film f
WHERE rating IN ('G', 'NC-17', 'PG-13', 'PG', 'R')
```

### d) Number of addresses in each district. (dplyr bonus)

```{sql connection=con_sakila}
SELECT district, COUNT(address_id)
FROM address
GROUP BY district
```
### e) Find the movies where rental rate is greater than $1 and order result set by descending order. (dplyr bonus)

### f) Top 2 movies that are rated R with the highest replacement cost. (dplyr bonus)

### g) Find the most frequently occurring (mode) rental rate across products. (dplyr bonus)

### h) Find the top 2 movies with movie length greater than 50mins and which has commentaries as a special features. (dplyr bonus)

### i) List the years with more than 2 movies released. (dplyr bonus)

# QUESTION 8 (Optional Practice Questions)

### a) Extract the street number (characters 1 through 4) from customer addressLine1. (dplyr bonus)

### b) Find out actors whose last name starts with character A, B or C. (dplyr bonus)

### c) Find film titles that contains exactly 10 characters. (dplyr bonus)

### d) Format a payment_date using the following format e.g "22/1/2016". (dplyr bonus)

### e) Find the number of days between two date values rental_date & return_date. (dplyr bonus)

